---
title: "Soil MLR with predictors normalised"
author: "Salman"
date: "2023-12-03"
output: html_document
---

Required libraries
```{r include=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(ggpmisc)
library(readxl)
library(caTools)
library(regclass)
library(glue)
library(tibble)
library(caret)
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

import the input file
```{r}
mlrinput <- read_excel("C:/soil-texture-map-french-guiana/data/field198_mlr.xlsx", sheet = 1)
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

normalising the in-situ observations used as predictors between 0 to 1 using min-max normlisation. mutate function from dplyr package

```{r}
normalize <- function(x) {
  if (is.data.frame(x)) {
    x %>% 
      mutate(across(everything(), ~ (.-min(.))/(max(.)-min(.)), .names = "norm_{.col}"))
  } else {
    (x - min(x)) / (max(x) - min(x))
  }}
```


```{r}
mlrinput <- mlrinput %>%
  mutate(across(10:16, ~ (.-min(.))/(max(.)-min(.)), .names = "norm_{.col}")) # min-max normalisation
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

creating the train control for k-fold random CV

```{r}
mlrctrl <- trainControl(method = "cv", number = 5) 
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++MLR model to predict sand

defining the covariates for sand model

```{r}
mlrsand_cov <- c("norm_rankcl", "norm_ranksi", "norm_ranksa", "norm_value", "norm_chroma", "norm_lith", "norm_cuirass")
```


sand content as a function of above field observations

```{r}
formula_mlrsand <- as.formula(paste("sand ~", paste(mlrsand_cov, collapse = " + ")))
```

building the sand model with 5 fold random CV

```{r}
mlrsand_model <- train(formula_mlrsand, data = mlrinput, method = "lmStepAIC", trControl = mlrctrl)
summary(mlrsand_model)
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++MLR model to predict clay, using observed clay as an add covariate

normalising sand values

```{r}
mlrinput$norm_sand <- normalize(mlrinput$sand)
```



defining the covariates for clay model

```{r}
mlrclay_cov <- c("norm_sand", "norm_rankcl", "norm_ranksi", "norm_ranksa", "norm_value", "norm_chroma", "norm_lith", "norm_cuirass")
```


sand content as a function of above field observations

```{r}
formula_mlrclay <- as.formula(paste("clay ~", paste(mlrclay_cov, collapse = " + ")))
```

building the sand model with 5 fold random CV

```{r}
mlrclay_model <- train(formula_mlrclay, data = mlrinput, method = "lmStepAIC", trControl = mlrctrl)
summary(mlrclay_model)
```

++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

developed MLR models to predict sand and clay at 198 points

```{r}
mlrinput$pred_sand <- predict(mlrsand_model, mlrinput)
```

```{r}
mlrinput$pred_clay <- predict(mlrclay_model, mlrinput)
```
correlations bw predicted and observed values of sand and clay
```{r}
cor(mlrinput$sand, mlrinput$pred_sand)
cor(mlrinput$clay, mlrinput$pred_clay)
```

++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

using the above developed models to predict sand and clay at 2810 (all in-situ observations)


loading the file

```{r}
insitu2810 <- read_excel("C:/soil-texture-map-french-guiana/data/field2810.xlsx", sheet = 1)
```
renaming col names

```{r}
insitu2810 <- insitu2810 %>%
  rename(IDSOIL = 1, SITE = 2)
```


omitting unused/unnecessary columns

```{r}
insitu2810_select <- insitu2810 %>%
  dplyr::select(IDSOIL, SITE, SONDAGE, lyr, rankcl, ranksi, ranksa, value, chroma, lith, cuirass)
```

normalising columns with values used as predictors

```{r}
insitu2810_select <- insitu2810_select %>%
  mutate(across(5:11, ~ (.-min(.))/(max(.)-min(.)), .names = "norm_{.col}")) # min-max normalisation
```

++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

predicting sand at 2810 points 

```{r}
insitu2810_select$pred_sand <- predict(mlrsand_model, insitu2810_select )
```

and to predict clay, we first normalise the (predicted) sand column


```{r}
insitu2810_select$norm_sand <- normalize(insitu2810_select$pred_sand)
```

predicting clay at 2810 points 

```{r}
insitu2810_select$pred_clay <- predict(mlrclay_model, insitu2810_select)
```

silt as the difference  ie 100 - sand - clay

```{r}
insitu2810_select$diff_silt <- 100 - insitu2810_select$pred_sand -insitu2810_select$pred_clay
```


++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

sand, silt and clay contents averaged across each soil profile

adding the depth of each layer to the df

```{r}
insitu2810_select <- insitu2810_select %>%
  mutate(lyr_dep = case_when(lyr == 1 ~ 100, lyr == 2 ~ 100, lyr == 3 ~ 200, lyr == 4 ~ 200, lyr == 5 ~ 200, lyr == 6 ~ 200, lyr == 7 ~ 200))
```

total depth of each soil profile

```{r}
insitu2810_select <- insitu2810_select %>%
  group_by(IDSOIL) %>%
  mutate(tot_dep = sum(lyr_dep, na.rm = TRUE))
```

relative proportion of each layer across the entire depth

```{r}
insitu2810_select$wt_lyr <- insitu2810_select$lyr_dep/ insitu2810_select$tot_dep
```

weighted average of sand, silt and clay in each soil profile

```{r}
avg_IDSOIL <- insitu2810_select %>%
  group_by(IDSOIL, SONDAGE, SITE) %>%
  summarise(avg_sand = weighted.mean(pred_sand, wt = wt_lyr, na.rm = TRUE), avg_silt = weighted.mean(diff_silt, wt = wt_lyr, na.rm = TRUE), avg_clay = weighted.mean(pred_clay, wt = wt_lyr, na.rm = TRUE))
```

+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
adding UTM coordinates to the dataset



```{r}
avg_IDSOIL_utm <- merge(avg_IDSOIL, insitu2810[, c("IDSOIL", "xutm22N", "yutm22N")], by = "IDSOIL", all.x = TRUE)
```

```{r}
avg_IDSOIL_utm <- avg_IDSOIL_utm[!duplicated(avg_IDSOIL_utm$IDSOIL), ]
```

```{r}
avg_IDSOIL_utm$xutm22N <- as.numeric(avg_IDSOIL_utm$xutm22N)
avg_IDSOIL_utm$yutm22N <- as.numeric(avg_IDSOIL_utm$yutm22N)
```


removing removes with missing lat/long values
```{r}
sol421_utm <- avg_IDSOIL_utm %>%
  filter(complete.cases(xutm22N, yutm22N))
```
saving the output

```{r}
write.csv(sol421_utm, "C:/soil-texture-map-french-guiana/outputs/sol421_utm.csv")
```


++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

